---
title: "Colorectal cancer: Patient characteristics and survival "
author: "Emily Wu"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

#package to help load packages
if (!require("pacman")) install.packages("pacman")

#install/load required packages
pacman::p_load(
  here,
  janitor,
  tidyverse,
  lubridate,
  survival,
  survminer,
  gtsummary,
  labelled
)

```

```{r download-data, include = FALSE}

#download data from the web to folder where this .Rmd file is located
# url <- "https://data.mendeley.com/public-files/datasets/vvzw3wkx93/files/bbb9836d-8f6e-473d-8b10-f55685cedc2a/file_downloaded"
# 
# myfile <- here("colorectal.csv")
# 
# download.file(url, myfile)
#colorectal <- read.csv(myfile, sep = ";")

#read in downloaded data 
myfile <- here("colorectal.csv")
colorectal <- read.csv(myfile, sep = ";")
```

```{r clean-data, include = FALSE}

#clean up data
crec <- colorectal %>% 
  
  #format variable names
  janitor::clean_names() %>%
  
  #variables I want to keep
  select(age, sex, m_status, edu_status, region,
        hist_type, location, tnm_stage, clinical_stage, cea_baseline, 
         chemotherapy, surgery, radiotherapy, 
         date_dx, date_death, vstatus) %>%
  
  #calculate time from diagnosis to death/censoring
  mutate(date_dx = lubridate::as_date(date_dx, format = "%m/%d/%Y"),
         date_death = lubridate::as_date(date_death, format = "%m/%d/%Y"),
         time = round(as.duration(date_dx %--% date_death) / dmonths(1))) %>%
  
  #create variable to describe overall treatment
  mutate(treatment_type = case_when(
    chemotherapy == 1 & surgery == 1 & radiotherapy == 1 ~ "Surgery and Chemotherapy and Radiotherapy",
    chemotherapy == 1 & surgery == 1 & radiotherapy == 0 ~ "Surgery and Chemotherapy only",
    chemotherapy == 1 & surgery == 0 & radiotherapy == 1 ~ "Chemotherapy and Radiotheraapy  only",
    chemotherapy == 1 & surgery == 0 & radiotherapy == 0 ~ "Chemotherapy only",
    chemotherapy == 0 & surgery == 1 & radiotherapy == 1 ~ "Surgery and Radiotheraapy  only",
    chemotherapy == 0 & surgery == 1 & radiotherapy == 0 ~ "Surgery only",
    chemotherapy == 0 & surgery == 0 & radiotherapy == 1 ~ "Radiotheraapy  only",
    chemotherapy == 0 & surgery == 0 & radiotherapy == 0 ~ "No treatment"
    ),
    
    treatment_cat = fct_collapse(factor(treatment_type), 
                                 no_surg = c("No treatment", "Chemotherapy only"),
                                 surg = c("Surgery only", "Surgery and Radiotheraapy  only"),
                                 chemo_plus_surg = c("Surgery and Chemotherapy only", 
                                                     "Surgery and Chemotherapy and Radiotherapy")) %>%
                                 fct_relevel("no_surg", "surg", "chemo_plus_surg")) %>%
  
  mutate(region = ifelse(region == "Addis Ababa", 1, 0)) %>%

  #label variables and values for better printed outputs
  set_variable_labels(
    age = "Age",
    sex = "Sex", 
    m_status = "Marital Status",
    edu_status = "Educational Status",
    region = "Residence Region",
    hist_type = "Histologic Type", 
    location = "Tumor Location",
    tnm_stage = "Tumor Stage",
    clinical_stage = "Clinical Stage",
    cea_baseline = "Baseline CEA", 
    chemotherapy = "Chemotherapy", 
    surgery = "Surgery", 
    radiotherapy = "Radiotherapy",
    treatment_type = "Treatment Received") %>%
  
  set_value_labels(
    sex = c(Male = 0, Female = 1),
    m_status = c(Single = 1, Married = 2, Widowed = 3, Divorced = 4),
    edu_status = c("No formal education" = 1, "Primary Level" = 2, "Secondary" = 3, "Higher education" = 4),
    region = c("Addis Ababa" = 1, "Out of Addis Ababa" = 0),
    location = c(Colon = 1, "Recto-sigmoid junction" = 2, Rectum = 3, Anorectal = 4),
    tnm_stage = c(T1 = 1, T2 = 2, T3 = 3, T4 = 4),
    clinical_stage = c("Localized" = 1, "Locally advanced" = 2, "Metastasis" = 3),
    hist_type = c("Adenocarcinoma NOS" = 1, "Mucinous/Signet-ring cell carcinoma" = 2),
    cea_baseline = c("Not Elevated (<5 ng/ml)" = 0, "Elevated (>=5 ng/ml)" = 1),
    chemotherapy = c(No = 0, Yes = 1),
    surgery  = c(No = 0, Yes = 1),
    radiotherapy = c(No = 0, Yes = 1)) %>%
   
   #convert certain variables to factors
   mutate_at(vars(c(sex, m_status, edu_status, region, hist_type, location, tnm_stage, clinical_stage,  cea_baseline, chemotherapy, surgery, radiotherapy)), to_factor) %>%
   
   #only 5 stage 1 cancers, so combine stage 1 and 2 cancers for later analysis
   mutate(tnm_stage_combined = fct_collapse(tnm_stage, "T1 and T2" = c("T1","T2")))
```

Colorectal cancer (CRC) is the third most common cancer worldwide. While colorectal cancer has primarily been a concern for highly developed countries, colorectal cancer incidence rates are now rising rapidly in low to middle income countries. 
Data on CRC in African countries, including Ethiopia, has been limited. A retrospective study from the Tikur Anbessa Specialized Hospital cancer registry provides an opportunity to better understand the characteristics, survival patterns, and prognostic factors of CRC patients in Ethiopia. 

The data comes from the medical records of 161 colorectal cancer patients registered at Tikur Anbessa Specialized Hospital, the primary oncology center in Ethiopia. Patients were followed up starting from Jan 1, 2012 to Dec 31, 2016. The follow-up time was from the first date of confirmed diagnosis to the date of death, date of loss to follow up or date of last contact.

### Table 1: Sociodemographic characteristics of patients
The mean age of patients was `r round(mean(crec$age))` years (SD: `r round(sd(crec$age))`). The majority of patients were male and married, and about half of patients resided in Addis Ababa. 

```{r make-table1}
table1 <- crec %>% 
  select(age, sex, edu_status, m_status, region) %>% 
  tbl_summary(                                             # stratify entire table by variable
    statistic = list(all_continuous() ~ "{mean} ({sd})",        # stats and format for continuous columns
                     all_categorical() ~ "{n} ({p}%)"),         # stats and format for categorical columns
    digits = all_continuous() ~ 1,                              # rounding for continuous columns
    type   = all_categorical(dichotomous = FALSE) ~ "categorical"                  # force all categorical levels to display
  ) 

table1
```

### Table 2: Pathological and clincial characteristics of patients
Out of 161 colorectal cancer cases, 156 (97%) were adenocarcinoma not otherwise specified. Colorectal cancer can occur in the colon, recto-sigmoid junction, rectum or anorectal regions. The majority of patients had cancer located in the colon. About half of patients had late tumor staged cancer. In additional, about half of patients had metastasized cancers. 
```{r make-table2}
table2 <- crec %>% 
  select(hist_type, location, tnm_stage, clinical_stage) %>% 
  tbl_summary(                                             # stratify entire table by variable
    statistic = list(all_continuous() ~ "{mean} ({sd})",        # stats and format for continuous columns
                     all_categorical() ~ "{n} ({p}%)"),         # stats and format for categorical columns
    digits = all_continuous() ~ 1,                              # rounding for continuous columns
    type   = all_categorical(dichotomous = FALSE) ~ "categorical"                  # force all categorical levels to display
  ) 

table2
```

### Table 3: Treatments received by patients
Patients received different combinations of surgery, chemotherapy and radiotherapy treatments. `r sum(crec$surgery == "Yes")` patients had surgery with or without adjunctive chemotherapy and radiotherapy. 

```{r make-table3}
table3 <- crec %>% 
  select(treatment_type) %>% 
  tbl_summary(                                             # stratify entire table by variable
    statistic = list(all_continuous() ~ "{mean} ({sd})",        # stats and format for continuous columns
                     all_categorical() ~ "{n} ({p}%)"),         # stats and format for categorical columns
    digits = all_continuous() ~ 1,                              # rounding for continuous columns
    type   = all_categorical(dichotomous = FALSE) ~ "categorical"                  # force all categorical levels to display
    
  ) 

table3
```

```{r examine-cea-prognostic, include = FALSE}

tab <- crec %>% 
  select(age, sex, tnm_stage, hist_type, cea_baseline, chemotherapy, surgery, radiotherapy) %>% 
  tbl_summary(     
    by = cea_baseline,                                          # stratify entire table by variable
    statistic = list(all_continuous() ~ "{mean} ({sd})",        # stats and format for continuous columns
                     all_categorical() ~ "{n} ({p}%)"),         # stats and format for categorical columns
    digits = all_continuous() ~ 1,                              # rounding for continuous columns
    type   = all_categorical(dichotomous = FALSE) ~ "categorical"                  # force all categorical levels to display
  ) %>%
   modify_spanning_header(c("stat_1", "stat_2") ~ "**Baseline CEA**") 

survminer::ggsurvplot(
    fit = survfit(Surv(time, vstatus) ~ cea_baseline, data = crec), 
    #risk.table = TRUE,
    xlab = "Months", 
    ylab = "Survival probability",
    legend.title = "Baseline CEA",
    legend.labs = c("<5 ng/ml", ">=5 ng/ml"))

```

### Survival pattern of patients
`r sum(crec$vstatus == 1)`(`r round(sum(crec$vstatus == 1)/nrow(crec)*100)`%) patients died due to colorectal cancer over a follow-up of at most 5 years. The median follow-up time was `r round(median(crec$time))` months. The overall colorectal cancer specific survival rate was low with median survival time of 21 months.  

```{r plot-km}
kmfit <- survfit(Surv(time, vstatus) ~ 1, data = crec)
survminer::ggsurvplot(
    fit = kmfit , 
    #risk.table = TRUE,
    xlab = "Months", 
    ylab = "Overall Survival probability")
  
```

### Proportional hazards model
A multivariable Cox proportional hazards was fit to the survival data. Stage 4 tumor, a histologic type of Mucinous/Signet-ring cell carcinoma, elevated baseline carcinoembryonic antigen (CEA) level, and surgical treatment were associated with worsened survival outcomes. 

```{r fit-cox}
model <- coxph(Surv(time, vstatus) ~ 
               age + sex + tnm_stage_combined + hist_type + cea_baseline + chemotherapy + surgery + radiotherapy,
               data = crec)

model %>%
  tbl_regression(exponentiate = TRUE)

```
